---
title: "R for Data Science chapter 3"
author: "Alibek Galiyev"
date: "September 7, 2018"
output:
    html_document:
        keep_md: true
---

# Chapter 3

```{r load_libraries}
library(nycflights13)
library(tidyverse)
```

To try _dplyr_ package we will use nyflights13::flights data, which is all flights departed from NY in 2013. 
Let's see data itself:

```{r data}
df = nycflights13::flights

head(df)
```

This data in tibble format, which is more optimized to work with tidyverse. All columns have their own 
types as int, dbl, chr, dttm. There's 7 types in tibble dataformat:

* int - integer numbers
* dbl - real numbers
* chr - character vectors or strings
* dttm - datetime
* lgl - boolean TRUE or FALSE
* fctr - categorical variables
* date - dates

## dplyr basics
dplyr package has 5 key functions:

* filter() - select given rows from dataset
* arrange() - reorder rows
* select() - pick columns by their names
* mutate() - create new columns
* summarize() - grouping data by some summary stats

## Filter Rows
*filter()* allows to subset observations based on some logic. For example, if we want to select all 
flights from January 1st:

```{r filter}
jan1 = filter(df, month == 1, day == 1)
```

R saves result to a variable. If you want to print out filtering and assign to some variable, you can wrap 
up into parentheses.

```{r filter_and_print}
(dec25 = filter(df, month == 12, day == 25))
```


## Comparisons
R has standard comparison operators like: >, <, >=, <=, ==. Since R store finite number of digits in float 
numbers, you can observe some strange behavior:

```{r comparisons}
sqrt(2) ^ 2 == 2
```

Instead of using "==", you should use _near()_ function for those kind of comparisons.

```{r near}
near(sqrt(2)^2, 2)
```


## Logical Operators
For "and" operator you can use "&" sign, for "or" use "|", and "!" for "not".

```{r logical}
filter(df, month == 11 | month == 12)
```

Instaed of using "|" in above exercise, it is more appropriate to use %in%.

```{r in}
nov_dec = filter(df, month %in% c(11, 12))
```


## Arrange Rows with arrange()
_arrange()_ works in similar way as filter, except it just reorder rows in given columns.

```{r arrange}
arrange(df, year, month, day)
```

_desc()_ function will make reverse order of values in columns:

```{r desc_arrange}
arrange(df, desc(arr_delay))
```

Important note, that missing vaslues (NAs) are always sorted at the end!


## Select Columns with select()
With _select()_ function you can select given columns by their names.

```{r select}
select(df, year, month, day)
```

Select all columns between year and day:
```{r select_year_day}
select(df, year:day)
```

Select all columns except from year and day:

```{r select_except}
select(df, -(year:day))
```

There are some helper function to use in select:

- starts_with("a") matches column names which starts with "a"
- ends_with("a") matches column names which ends with "a"
- contains("dgh") matches column names which contains "dgh"
- matches("(.)\\1") matches column names due to regular expression
- num_range("col", 1:3) matches "col1", "col2", "col3"

To rename column name you should use _rename()_ function:

```{r rename}
rename(df, tail_num = tailnum)
```

If you want to move some columns to the begining, function _everything()_ can be helpful:

```{r everything}
select(df, time_hour, air_time, everything())
```













